### 2010-10-06 Documentation/atropos/atropos.tex
made some points about significant contributions
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@651 fd2da0a4-d968-48df-9cb5-d8c496287cbf

% Set page margins correctly
\usepackage{geometry}
\geometry{letterpaper,top=1.0in,left=1.0in,bottom=1.0in,top=1.0in,headsep=6pt,footskip=18pt}

@@

While Atropos may be applied to multivariate data from arbitrary
modalities, we limit our Atropos evaluation to tissue-classification
in T1 neuroimaging in part due to the abundance of ``gold-standard''
data for this modality.  Consistent with our advocacy of open science
(not to mention the facilitation of analysis due to accessibility) we
also only use publicly available data sets.  Organization of this work
is as follows: we first describe the theory behind the various
algorithmic components of Atropos while acknowledging that theoretical
discussion is available elsewhere, e.g. in previously cited
literature.  This is followed by a thorough discussion of
implementation which is something not typically discussed but yet is
of immense practical utility.  Finally, we provide a discussion of our work both in the
context of the remainder of this paper and in the larger context of
previous and future work.

%\begin{comment}{

@@

Atropos encodes a family of Bayesian segmentation techniques that may be configured in an application-specific manner.  The theory underlying Atropos dates back 20+ years and is representative of some of the most innovative work in the field.  Although we synthesize some of the theoretical work in this section, we recommend to the interested reader consultation with the literature cited previously for further insight.

@@

@@

@@

\section{Comparative Evaluation}

{\color{red}{\em Brian, John}}

\subsection{The LPBA40 Dataset}  
\label{sec:lpba}
The LONI Probabilistic Brain Atlas data, LPBA40, \citep{Shattuck2008} was collected at the North Shore Long
Island Jewish Health System imaging center and is maintained at UCLA.
LPBA40 contains 40 images (20 male $+$ 20 female) from normal, healthy
ethnically diverse volunteers with average age of 29.2 $\pm$ 6.3
years.  Each subject underwent 3D SPGR MRI on a 1.5T GE system
resulting in $0.86 \times 0.86 \times 1.5 mm^3$ images.  Each MRI in
the LPBA40 dataset was manually labeled with 56 independent structures
at the UCLA Laboratory of Neuro Imaging (LONI).  The test-retest
reliability of the labeling, across raters, was reported as a minimum
Jaccard ratio of $0.697$ in the supramarginal gyrus to a maximum of
$0.966$ in the gyrus rectus.  A single labeling of each image is made
available to the public and used, here, as silver-standard data 
for both training and testing in our cross-validation scheme.

\subsection{The NIREP Dataset}  
The non-rigid image registration evaluation project (NIREP
http://www.nirep.org/) is a resource of 16 high quality labeled brain
images at 1$mm^3$.  Each brain was labeled with 32 cortical regions
(16 on each hemisphere) and an additional class of other gray matter
tissue.  Regions vary in size from large (inferior temporal region) to
small (temporal pole, insula gyrus, frontal pole).  The main drawback
is a lack of inter-rater reliability numbers -- in particular because
visual inspection and comparison of labelings reveals a degree of
inconsistency in labeling of particular regions across subjects.
Nevertheless, the NIREP dataset is perhaps the highest quality
evaluation dataset currently available for the cortex.

>>>>

% Set page margins correctly
\usepackage{geometry}
\usepackage{url}
\geometry{letterpaper,top=1.0in,left=1.0in,bottom=1.0in,top=1.0in,headsep=6pt,footskip=18pt}

@@

The major contribution of the Atropos algorithm is a flexible, memory
efficient implementation of core expectation maximization machinery
that is useful across many segmentation problems.  To highlight the
value of this open-source, free contribution, we performed a search of
software attributes on NITRC and found that as of {\textcolor{red}
INSERT DATE NO STAND-ALONE N-CLASS EM methods are currently listed}.
Atropos's memory efficiency enables EM-segmentation of many classes at
once, a capability that is due to a sparse internal representation of
prior probability images that minimizes the need for random access
memory.  We evaluate EM-based classification of $1 mm^3$ T1 MR
neuroimages into 68 neuronatomical regions to illustrate the practical
value of the low-memory implementation within this paper.  Atropos
also allows multivariate image features to drive the data term.  Such
an approach is valuable when more than one view of anatomy aids
segmentation, as in neonatal brain tissue classification
\cite{Weisenfeld2009}.  Although Atropos may be applied to
multivariate data from arbitrary modalities, we limit our evaluation
to tissue-classification in T1 neuroimaging in part due to the
abundance of ``gold-standard'' data for this modality.  Consistent
with our advocacy of open science (not to mention the facilitation of
analysis due to accessibility) we also only use publicly available
data sets.

Organization of this work is as follows: we first describe the theory
behind the various algorithmic components of Atropos while
acknowledging that theoretical discussion is available elsewhere,
e.g. in previously cited literature.  This is followed by a thorough
discussion of implementation which, though often overlooked, is of
immense practical utility.  We then report evaluation results on
public T1 neuroimaging datasets that have reference gold/silver
standard labelings.  Finally, we provide a discussion of our results
and our open-source contribution in the context of the remainder of
this paper and of previous and future work.

@@

Atropos encodes a family of Bayesian segmentation techniques that may be configured in an application-specific manner.  The theory underlying Atropos dates back 20$+$ years and is representative of some of the most innovative work in the field.  Although we synthesize some of the theoretical work in this section, we recommend to the interested reader consultation with the literature cited previously for further insight.

@@

Figure~\ref{fig:neonate} illustrates the benefit of geodesic distance
priors in segmenting T2 MRI of the 
neonatal brain.  Geodesic distance priors may be used to reduce ambiguity at
tissue interfaces, where partial voluming leads to mixed intensity.  
In neonatal brain segmentation from T2 MRI, the
interface of gray matter (low intensity) and cerebrospinal fluid
(bright intensity) often matches that of white matter (medium intensity). 
However, the white matter class should never appear near the surface
of the brain.  Thus, the distance from the brain surface may be used
to disambiguate voxels such as these.  
In this case, the distance of a voxel from a known
or expected tissue location serves to differentiate it from classes
that have similar intensity signature.  The geodesic distance is therefore an
alternative to explicit models of partial volume \citep{Ruan2000,Ballester2002,Leemput2003}.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.9\textwidth]{sparse_labels.png} 
\caption{\baselineskip 12pt \small  Neonatal brain segmentation.  }
\label{fig:neonate}
\end{center}
\end{figure}

@@

\section{Evaluation}

\input{evaluation_data_description.tex}
